---
title: "Haplotype analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(plotly)
library(cowplot)
library(vegan)
library(UpSetR)
library(ggpubr)
library(ggplot2); theme_set(theme_bw()); theme_update(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

sample.colors = c(`mother` = "#E2BD4D"  , `0m`= "#27A4C3" , `1m`= "#004D40",`3m`="#E25D8D", `9m`="#7152A4")

group.colors <- c(`Haplotype_0` = "#7BCEC8", 
                  `Haplotype_1` = "#FF9896", 
                  `Haplotype_2` ="#B4E877", 
                  `Haplotype_3` = "#FFDC73",
                  `Haplotype_4` = "#5DA069",
                  `Haplotype_5` = "#6291EA",
                  `Haplotype_6` = "#C5B0D5",
                  `Haplotype_7` = "#C49C94", 
                  `Haplotype_8` = "#D46A97", 
                  `Haplotype_9` ="#FFB14D", 
                  `Haplotype_10` = "#DE9ED6",
                  `Haplotype_11` = "#6B6ECF")


```


Read in seqtab, output from dada2 script
```{r}
#read in seqtab and do some pruning 

#forward
seqtab.nochim_fw = read.csv("../dada2/out/seqtab_nochim_fw.csv")

#remove ASV with population freq lower than .1%
# find sum of the matrix
Matrix_sum = seqtab.nochim_fw[,2:ncol(seqtab.nochim_fw)] %>% 
  sum()
# Find the number of reads that would be less that 1% across population
cutoff_reads = Matrix_sum * 0.001

seqtab_prune_fw = seqtab.nochim_fw %>% as.data.frame() %>% 
  select_if(negate(function(col) is.numeric(col) && sum(col) < cutoff_reads)) %>% 
  rename(sample  = `X`) 

seqtab_prune_fw %>% dim()
seqtab.nochim_fw %>% dim

seqtab.nochim_fw %>% 
  column_to_rownames("X") %>% 
  pivot_longer(names_to = "seq", values_to = "read", cols =everything()) %>% 
  ggplot()+geom_histogram(aes(x=read))+scale_x_log10() +geom_vline(xintercept = cutoff_reads)

#write.csv(seqtab_prune, "../dada2/out/seqtab_prune_fw.csv", row.names = F)
#seqtab_prune = read.csv("../dada2/out/seqtab_prune_fw.csv") 


seqtab.nochim_rv = read.csv("../dada2/out/seqtab_nochim_rv.csv")

#remove ASV with population freq lower than .1%
# find sum of the matrix
Matrix_sum = seqtab.nochim_rv[,2:ncol(seqtab.nochim_rv)] %>% 
  sum()
# Find the number of reads that would be less that 1% across population
cutoff_reads = Matrix_sum * 0.001

seqtab_prune_rv = seqtab.nochim_rv %>% as.data.frame() %>% 
  select_if(negate(function(col) is.numeric(col) && sum(col) < cutoff_reads)) %>% 
  rename(sample  = `X`) 

seqtab_prune_rv %>% dim()
seqtab.nochim_rv %>% dim


seqtab.nochim_rv %>% 
  column_to_rownames("X") %>% 
  pivot_longer(names_to = "seq", values_to = "read", cols =everything()) %>% 
  ggplot()+geom_histogram(aes(x=read))+scale_x_log10()+geom_vline(xintercept = cutoff_reads)

```

# Generate long dataformats
```{r}

seqtab_work_fw = seqtab_prune_fw %>% 
  pivot_longer(cols = -sample, names_to = "seq", values_to = "freq")
seqtab_work_rv = seqtab_prune_rv %>% 
  pivot_longer(cols = -sample, names_to = "seq", values_to = "freq")


#merge fw and rv by ASV
seqtab_work = full_join(seqtab_work_fw, seqtab_work_rv, by = c("seq", "freq", "sample") ) %>% 
  mutate(source =ifelse(str_detect(sample, "fw"), "fw","rv")) %>% 
  mutate(samples_merged = substring(sample, 1, nchar(sample) -2 )) %>% 
  mutate(MID = str_extract(sample, "(?<=MID_)[:digit:]")) 


```



# Output sequences 
```{r eval=FALSE, include=FALSE}
#output the sequences to do some clustering 
k = 0
for (i in (seqtab_work_fw$seq %>% unique())) {
    cat(">", "Putative_Haplotype_", k, "\n", i, "\n") 
    k = k + 1
} 

```

MANUALLY CHANGED hap 6 to hap 12


# Make seqtab annotadated
```{r}
seqtab_annotated = seqtab_cluster %>% 
  group_by(samples_merged, seq) %>% 
  mutate(merged_freq = sum(freq)) %>%
  filter(merged_freq!=0) %>% 
  select(-source, -sample, -freq) %>%
  unique() %>% 
  ungroup() %>% 
  mutate(mother = ifelse(str_detect(samples_merged, "mother"), "mother", "child")) %>% 
  mutate(samples_merged = case_when(samples_merged == "35_11m_MID_1" ~ "035_11m_MID_1",
                            samples_merged == "35_11m_MID_2" ~ "035_11m_MID_2",
                            samples_merged == "35_9m_MID_1"~"035_9m_MID_1", 
                            samples_merged == "35_9m_MID_2"~"035_9m_MID_2",
                            TRUE ~samples_merged
                            )) %>% 
  mutate(sample_type = str_extract(samples_merged, "[:alnum:]{2,6}(?=_MID)")) %>% 
  mutate(family = str_extract(samples_merged, ".{3,4}(?=_)")) %>% 
  mutate(MID = str_extract(samples_merged, "(?<=MID_)[:digit:]")) %>% 
  mutate(sample_single = str_extract(samples_merged, ".{4,11}(?=_MID_)")) %>% 
  group_by(samples_merged) %>% 
  mutate(total_reads = sum(merged_freq)) %>% 
  group_by(community_cluster,samples_merged) %>% 
  mutate(cluster_freq = sum(merged_freq)) %>% 
  ungroup()


#write.csv(seqtab_annotated, "../dada2/out/seqtab_annotated.csv", row.names = F)

seqtab_annotated = read.csv("../dada2/out/seqtab_annotated.csv") %>% 
  mutate(seq_id = ifelse(seq_id == "Haplotype_12", "Haplotype_0", seq_id)) %>% 
  select(-community_cluster, -cluster_freq, -seq) %>% 
  group_by(seq_id, samples_merged) %>% mutate(merged_freq = sum(merged_freq)) %>% ungroup() %>% unique() %>% 
  mutate(family =as.character(ifelse(family == "35", "035", family))) %>% 
  mutate(sample_type = ordered(sample_type, 
                           levels = c("mother", "0m", "1m","3m","9m"))) %>% 
  mutate(family = ordered(family, 
                          levels =c("117", "4296", "4825", #twins
                                    "4285","3482","3236","2134","4186", #have mom, two reps
                                    "1244","1507","2315","2619","2596", #have mom, one rep
                                    "3656","1559","035","500","2838","3048","3103","3565", #No mom, two reps/timepoints
                                    "869","181","327") )) %>% 
  ungroup()
```



# How abundant is each ASV in the whole population
```{r}
seqtab_annotated %>% 
  group_by(seq_id) %>% 
  summarise(seq_id_freq = sum(merged_freq)) %>% 
  ggplot(aes(x = seq_id, y = seq_id_freq)) +
  geom_col() +
  geom_text(aes(label = seq_id_freq), vjust = -.35)+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```



# Ascertaining the cutoff of total reads > 200 for sample inclusion
Compare read depth for each sample to shannons diversity and total number of haplotypes recovered
```{r}

# Using Shannon

seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  pivot_wider(id_cols = c(samples_merged), names_from =  seq_id, values_from = merged_freq, values_fn =  sum ) %>% 
  replace(is.na(.), 0) %>% 
  column_to_rownames("samples_merged") %>% 
  diversity(index= "shannon") %>% 
  as.matrix() %>% as.data.frame() %>% 
  rownames_to_column("samples_merged") %>% rename(shannon = V1) %>% 
  mutate(sample_type = str_extract(samples_merged, "[:alnum:]{2,6}(?=_MID)")) %>% 
  mutate(family = str_extract(samples_merged, ".{3,4}(?=_)")) %>% 
  mutate(MID = str_extract(samples_merged, "(?<=MID_)[:digit:]")) %>% 
    mutate(sample_single = str_extract(samples_merged, ".{4,11}(?=_MID_)")) %>% 
  group_by(sample_single) %>% 
  mutate(mean_shannon = (mean(shannon))) %>% 
  left_join(seqtab_annotated %>% select(samples_merged, total_reads), by = "samples_merged") %>% 
  ggplot(aes(y = shannon, x = total_reads,color = sample_type))+
  scale_color_manual(values = sample.colors)+
  scale_x_log10()+
  ylab("Shannon") +xlab("Total Sample Reads")+
  geom_point( size = 3) 



#Using raw number of ASVs
seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  ungroup() %>% 
  group_by(samples_merged) %>% 
  mutate(num_haplotypes = n()) %>% 
  ggplot(aes(x=total_reads, y= num_haplotypes,color = sample_type )) +
  scale_color_manual(values = sample.colors)+
  geom_point(aes( text= samples_merged), size = 3) +
  ylab("Number of Haplotypes") +xlab("Total Sample Reads")+
  scale_x_log10() 

## calc correlations
## overall
temp = seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  ungroup() %>% 
  group_by(samples_merged) %>% 
  mutate(num_haplotypes = n()) 

cor.test(temp$num_haplotypes, temp$total_reads, method = "spearman")
# moms
temp1 = temp %>% filter(sample_type == "mother")
cor.test(temp1$num_haplotypes, temp1$total_reads, method = "spearman")

#0m
temp1 = temp %>% filter(sample_type == "0m")
cor.test(temp1$num_haplotypes, temp1$total_reads, method = "spearman")


#overall
temp = seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  pivot_wider(id_cols = c(samples_merged), names_from =  seq_id, values_from = merged_freq, values_fn =  sum ) %>% 
  replace(is.na(.), 0) %>% 
  column_to_rownames("samples_merged") %>% 
  diversity(index= "shannon") %>% 
  as.matrix() %>% as.data.frame() %>% 
  rownames_to_column("samples_merged") %>% rename(shannon = V1) %>% 
  mutate(sample_type = str_extract(samples_merged, "[:alnum:]{2,6}(?=_MID)")) %>% 
  mutate(family = str_extract(samples_merged, ".{3,4}(?=_)")) %>% 
  mutate(sample_single = str_extract(samples_merged, ".{4,11}(?=_MID_)")) %>% 
  group_by(sample_single) %>% 
  mutate(mean_shannon = (mean(shannon))) %>% 
  left_join(seqtab_annotated %>% select(samples_merged, total_reads), by = "samples_merged") 

cor.test(temp$mean_shannon, temp$total_reads, method = "spearman")
```


# Make family plots 
stacked bar charts, normalized and not normalized, describing haplotype composition for each sample, grouped by family.

```{r}
Makeplot <- function(fam){
    seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == fam) %>% 
    ggplot()+
    scale_fill_manual(values = group.colors)+
    geom_col(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black") +
    theme(legend.position = "none")+
    facet_wrap(~family) 
}
pdf("../Amplicon_analysis/Figures/ASV_byfam.pdf", height = 20, width = 24)
plot_grid(Makeplot("035"), 
          Makeplot("117"),
          Makeplot("181"),
          Makeplot("1244"),
          Makeplot("1507"),
          Makeplot("1559"),
          Makeplot("2134"),
          Makeplot("2315"),
          Makeplot("2596"),
          Makeplot("2619"),
          Makeplot("2838"),
          #Makeplot("2898"),
          Makeplot("3048"),
          Makeplot("3103"),
          Makeplot("3236"),
          Makeplot("327"),
          Makeplot("3482"),
          Makeplot("3565"),
          Makeplot("3656"),
          Makeplot("4186"),
          Makeplot("4285"),
          Makeplot("4296"),
          #Makeplot("4533"),
          #Makeplot("4635"),
          #Makeplot("4930"),
          Makeplot("500"),
          Makeplot("869")
          )
dev.off()

Makeplot <- function(fam){
    seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == fam) %>% 
    ggplot()+
    geom_bar(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black", position = "fill", stat = "identity") +
    scale_fill_manual(values = group.colors)+
    theme(legend.position = "none")+
    facet_wrap(~family) 
}

pdf("../Amplicon_analysis/Figures/ASV_byfam_read_norm.pdf", height = 20, width = 24)
plot_grid(Makeplot("035"), 
          Makeplot("117"),
          Makeplot("181"),
          Makeplot("1244"),
          Makeplot("1507"),
          Makeplot("1559"),
          Makeplot("2134"),
          Makeplot("2315"),
          Makeplot("2596"),
          Makeplot("2619"),
          Makeplot("2838"),
          #Makeplot("2898"),
          Makeplot("3048"),
          Makeplot("3103"),
          Makeplot("3236"),
          Makeplot("327"),
          Makeplot("3482"),
          Makeplot("3565"),
          Makeplot("3656"),
          Makeplot("4186"),
          Makeplot("4285"),
          Makeplot("4296"),
          #Makeplot("4533"),
          #Makeplot("4635"),
          #Makeplot("4930"),
          Makeplot("500"),
          Makeplot("869")
          )
dev.off()
```

Generate induvidual plots highlighted in main figure 3
```{r}
seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == "4296") %>% 
    ggplot()+
    geom_bar(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black", position = "fill", stat = "identity") +
    scale_fill_manual(values = group.colors)+
    xlab("Normalized Reads")+
    facet_wrap(~family) 

seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == "1507") %>% 
    ggplot()+
    geom_bar(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black", position = "fill", stat = "identity") +
    scale_fill_manual(values = group.colors)+
    xlab("Normalized Reads")+
    facet_wrap(~family) 

seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == "2619") %>% 
    ggplot()+
    geom_bar(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black", position = "fill", stat = "identity") +
    scale_fill_manual(values = group.colors)+
    xlab("Normalized Reads")+
    facet_wrap(~family) 

seqtab_annotated %>% 
    filter(total_reads>200) %>% 
    filter(family == "3236") %>% 
    ggplot()+
    geom_bar(aes(x = merged_freq, y = samples_merged, fill = seq_id),color = "black", position = "fill", stat = "identity") +
    scale_fill_manual(values = group.colors)+
    xlab("Normalized Reads")+
    facet_wrap(~family) 
```


# Haplotype frequency heatmap 
supplemental figure 1
```{r}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}


#Annotation information

#make num of haplotypes
test = seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  select(samples_merged, seq_id) %>% 
  group_by(samples_merged) %>% 
  mutate(num_haps = n()) %>% 
  left_join(seqtab_annotated, by = c("samples_merged", "seq_id")) 

annotation = data.frame(samples_merged = test$samples_merged,
                        #family = seqtab_annotated$family,
                        #replicate = as.character(test$MID),
                        number_haps = test$num_haps,
                        sample_type = test$sample_type
                        ) 

annotation = annotation  %>% 
  unique() %>%
  remove_rownames() %>% 
  column_to_rownames(var = "samples_merged")

## Raw ASV
seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  filter(sample_type !=4186) %>% 
  group_by(samples_merged) %>% 
  mutate(reads_per_sample = sum(merged_freq)) %>% 
  mutate(psuedocount = ifelse(merged_freq==0, 1, merged_freq)) %>% 
  mutate(norm_reads = psuedocount/reads_per_sample) %>% 
  na.omit() %>% 
  group_by(seq_id, samples_merged) %>% 
  mutate(asv_freq_sum = sum(norm_reads)) %>% 
  arrange(family, sample_type) %>% 
  select(asv_freq_sum, samples_merged, seq_id) %>% unique() %>% na.omit() %>% 
  pivot_wider(id_cols = c("samples_merged","family", "sample_type"), names_from =seq_id , values_from = asv_freq_sum) %>% view
  select(-family, -sample_type) %>% 
  column_to_rownames("samples_merged") ->test 

test[is.na(test)] <- 0

test %>% 
  na.omit() %>% 
  pheatmap::pheatmap(
                    color = colorRampPalette(c( "white", "navy"))(40),
                    cluster_rows = F,
                    annotation_row = annotation,
                    border_color = "black",
                    annotation_colors = list(sample_type = 
                                             c(`mother` = "#E2BD4D"  , `0m`= "#27A4C3" , `1m`= "#004D40",`3m`="#E25D8D", `9m`="#7152A4"))
              )->p

pdf("../Amplicon_analysis/Figures/heatmapfrequency_ASV_readnorm.pdf", height = 16, width = 10)
p
dev.off()

## Raw ASV binary

seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  filter(sample_type !=4186) %>% 
  na.omit() %>% 
  group_by(seq_id, samples_merged) %>% 
  #select(merged_freq, samples_merged, seq_id) %>% unique() %>% na.omit() %>% 
  pivot_wider(id_cols = samples_merged, names_from =seq_id , values_from = merged_freq, values_fn = sum) %>%#fix values fn once u fixed 035
  pivot_longer(cols = -samples_merged, names_to = "seq_id", values_to = "merged_freq") %>% 
  mutate(hap_bin = ifelse(merged_freq !=0, 1, 0)) %>% 
  mutate(sample_type = str_extract(samples_merged, "[:alnum:]{2,6}(?=_MID)")) %>% 
  mutate(family = str_extract(samples_merged, ".{3,4}(?=_)")) %>% 
  pivot_wider(id_cols = c("samples_merged","family","sample_type"), names_from = seq_id, values_from = hap_bin) %>% 
  mutate(sample_type = ordered(sample_type, 
                           levels = c("mother", "0m", "1m","3m","9m"))) %>% 
  mutate(family = ordered(family, 
                          levels =c("117", "4296", "4825", #twins
                                    "4285","3482","3236","2134","4186", #have mom, two reps
                                    "1244","1507","2315","2619","2596", #have mom, one rep
                                    "3656","1559","035","500","2838","3048","3103","3565", #No mom, two reps/timepoints
                                    "869","181","327") )) %>% 
  arrange(family, sample_type) %>% select(-family,-sample_type) %>% 
  column_to_rownames("samples_merged") ->test 

test[is.na(test)] <- 0

test %>% 
  na.omit() %>% 
  pheatmap::pheatmap(
                    color = colorRampPalette(c( "white", "navy"))(10),
                    cluster_rows = F,
                    annotation_row = annotation,
                    border_color = "black",
                    annotation_colors = list(sample_type = 
                                             c(`mother` = "#E2BD4D"  , `0m`= "#27A4C3" , `1m`= "#004D40",`3m`="#E25D8D", `9m`="#7152A4"))
              )->p


pdf("../Amplicon_analysis/Figures/heatmapfrequency_ASV_readnorm_binary.pdf", height = 16, width = 10)
p
dev.off()


```


# Shannons in baby at 0 months vs mom
figure 3a
```{r}
##ggpubr


#All of them
ggviolin(test %>%
              select(sample_type,mean_shannon, sample_single) %>% unique(),
            x = "sample_type", y = "mean_shannon", fill = "sample_type",alpha = 0.7,
            desc_stat = "mean_ci", color = "black", palette = sample.colors, trim = T,
            add = c("jitter"), add.params = list(fill = "sample_type", alpha = 0.7) , xlab = "Sample Type", ylab = "Mean Shannon"
            )+
  stat_compare_means(comparisons = my_comparisons,
                     label = "p.signif",
                     method = "wilcox.test")
my_comparisons <- list( c("mother", "0m"))

ggviolin(test %>% filter(sample_type %in% c("0m", "mother")),
            x = "sample_type", y = "mean_shannon", fill = "sample_type",alpha = 0.7,
            desc_stat = "mean_ci", color = "black", palette = sample.colors, trim = T,
            add = c("jitter"), add.params = list(fill = "sample_type", alpha = 0.7) 
            )+
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "wilcox.test")


```


# Sequence differences: ASVs
figure 1a
```{r}
#Read in matrix from geneious

pdf("../Amplicon_analysis/Figures/heatmap_PID.pdf", height = 7, width = 8)
read.csv("../Amplicon_analysis/sequences/Distance_PID_ASV.csv") %>% 
  column_to_rownames("X") %>% 
  pheatmap::pheatmap(display_numbers = T)
dev.off()

p

```



# generate dataframes that will be used for figures that require "counting" of haplotypes
```{r}
both_samples = seqtab_annotated %>% 
  pivot_wider(id_cols = c("sample_single", "seq_id"), names_from = MID, values_from = merged_freq, values_fn = sum, values_fill = 0) %>% 
  mutate(has_hap_both = ifelse(`1` !=0 & `2` !=0, TRUE, FALSE)) %>% 
  mutate(has_hap_either = ifelse (`1` !=0 | `2` !=0, TRUE, FALSE) ) %>% 
  pivot_longer(cols = c(`1`,`2`), names_to = "MID", values_to = "merged_freq") %>% 
  mutate(MID = as.integer(MID)) %>% 
  left_join(seqtab_annotated, by = c("sample_single", "MID", "merged_freq", "seq_id")) %>% 
  filter(has_hap_both ==TRUE)


both_norep = seqtab_annotated %>% 
  filter(total_reads>200)  %>% 
  ungroup() %>% group_by(sample_single, seq_id) %>% 
  mutate(sum_mid = sum(MID)) %>%
  mutate(has_rep = ifelse(sum_mid %in% c("1","2"), "FALSE", "TRUE")) %>% 
  pivot_wider(id_cols = c("sample_single", "seq_id", "has_rep"), names_from = MID, values_from = merged_freq, values_fn = sum, values_fill = 0) %>% 
  mutate(has_hap_both = ifelse(`1` !=0 & `2` !=0, TRUE, FALSE)) %>% 
  mutate(has_hap_either = ifelse (`1` !=0 | `2` !=0, TRUE, FALSE) ) %>% 
  pivot_longer(cols = c(`1`,`2`), names_to = "MID", values_to = "merged_freq") %>% 
  mutate(MID = as.integer(MID)) %>% 
  left_join(seqtab_annotated, by = c("sample_single", "MID", "merged_freq", "seq_id")) %>% 
  filter(has_hap_both ==TRUE | has_rep == FALSE )


```

# how many haplotypes per sample
```{r}
#How many total samples
seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  select(sample_single) %>% unique() ->test1

# how many mothers and children
seqtab_annotated %>% 
  filter(total_reads>200) %>% 
  select( sample_single, sample_type) %>% unique() %>% select(sample_type) %>% 
  table()


both_norep %>% 
  select(samples_merged, seq_id, sample_single, merged_freq, sample_type) %>% 
  group_by(samples_merged) %>% 
  mutate(num_haps = n()) %>% 
  mutate(more_than_2 = ifelse(num_haps>2, TRUE,FALSE)) %>% 
  na.omit() %>% 
  ungroup() %>% group_by(sample_single) %>% 
  mutate(MID = str_extract(samples_merged, "(?<=MID_)[:digit:]")) %>% 
  #select(sample_single, more_than_2, sample_type) %>% unique() %>% view()
  select( sample_single, sample_type, more_than_2) %>% ungroup() %>%  unique() %>% filter(more_than_2 =="FALSE") %>% select(sample_type) %>% table()

union(test1$sample_single, test2$sample_single)
test2$sample_single
```



#upset plot
figure 1c

```{r}

## haps appear in both
C0 = both_samples %>%  filter(seq_id == "Haplotype_0")
C1 =both_samples %>% filter(seq_id == "Haplotype_1")
C2 =both_samples%>%  filter(seq_id == "Haplotype_2")
C3 =both_samples%>%  filter(seq_id == "Haplotype_3")
C4 =both_samples%>%  filter(seq_id == "Haplotype_4")
C5 =both_samples%>% filter(seq_id == "Haplotype_5")
C6 =both_samples%>%  filter(seq_id == "Haplotype_6")
C7 =both_samples%>%  filter(seq_id == "Haplotype_7")
C8 =both_samples%>%  filter(seq_id == "Haplotype_8")
C9 =both_samples%>%  filter(seq_id == "Haplotype_9")
C10 =both_samples%>%  filter(seq_id == "Haplotype_10")
C11=both_samples%>%  filter(seq_id == "Haplotype_11")
#C12=both_samples%>%  filter(seq_id == "Haplotype_12")

upset_list = list(Haplotype_0 =C0$sample_single %>% unique(),
                  Haplotype_1 = C1$sample_single %>% unique(),
                  Haplotype_2 = C2$sample_single %>% unique(),
                  Haplotype_3 = C3$sample_single %>% unique(),
                  Haplotype_4 = C4$sample_single %>% unique(),
                  Haplotype_5 = C5$sample_single %>% unique(),
                  Haplotype_6 = C6$sample_single %>% unique(),
                  Haplotype_7 = C7$sample_single %>% unique(),
                  Haplotype_8 = C8$sample_single %>% unique(),
                  Haplotype_9 = C9$sample_single %>% unique(),
                  Haplotype_10 = C10$sample_single %>% unique(),
                  Haplotype_11 = C11$sample_single %>% unique()
                  )

#make group that have each set to make a colored bar


pdf("../Amplicon_analysis/Figures/Upset_bothsamples.pdf", height = 5.5, width = 5)
upset(fromList(upset_list),
              order.by = "freq",
              nsets = length(upset_list),
              sets.bar.color = c(
                  `Haplotype_0` = "#7BCEC8", 
                  `Haplotype_6` = "#C5B0D5",
                  `Haplotype_7` = "#C49C94", 
                  `Haplotype_1` = "#FF9896", 
                  `Haplotype_4` = "#5DA069",
                  `Haplotype_9` ="#FFB14D",
                  `Haplotype_5` = "#6291EA",
                  #`Haplotype_12` = "#FFDC73",
                  `Haplotype_3` = "#17BECF",
                  `Haplotype_8` = "#D46A97", 
                  `Haplotype_2` ="#B4E877", 
                  `Haplotype_11` = "#6B6ECF",
                  `Haplotype_10` = "#DE9ED6"
                  ))
dev.off()

## haps appear in both OR there's only one rep
C0 = both_norep %>%  filter(seq_id == "Haplotype_0")
C1 =both_norep %>% filter(seq_id == "Haplotype_1")
C2 =both_norep%>%  filter(seq_id == "Haplotype_2")
C3 =both_norep%>%  filter(seq_id == "Haplotype_3")
C4 =both_norep%>%  filter(seq_id == "Haplotype_4")
C5 =both_norep%>% filter(seq_id == "Haplotype_5")
C6 =both_norep%>%  filter(seq_id == "Haplotype_6")
C7 =both_norep%>%  filter(seq_id == "Haplotype_7")
C8 =both_norep%>%  filter(seq_id == "Haplotype_8")
C9 =both_norep%>%  filter(seq_id == "Haplotype_9")
C10 =both_norep%>%  filter(seq_id == "Haplotype_10")
C11=both_norep%>%  filter(seq_id == "Haplotype_11")
C12=both_norep%>%  filter(seq_id == "Haplotype_12")

upset_list = list(Haplotype_0 =C0$sample_single %>% unique(),
                  Haplotype_1 = C1$sample_single %>% unique(),
                  Haplotype_2 = C2$sample_single %>% unique(),
                  Haplotype_3 = C3$sample_single %>% unique(),
                  Haplotype_4 = C4$sample_single %>% unique(),
                  Haplotype_5 = C5$sample_single %>% unique(),
                  Haplotype_6 = C6$sample_single %>% unique(),
                  Haplotype_7 = C7$sample_single %>% unique(),
                  Haplotype_8 = C8$sample_single %>% unique(),
                  #Haplotype_12 = C12$sample_single %>% unique(),
                  Haplotype_9 = C9$sample_single %>% unique(),
                  #Haplotype_10 = C10$sample_single %>% unique(),
                  Haplotype_11 = C11$sample_single %>% unique()
                  )

#make group that have each set to make a colored bar


pdf("../Amplicon_analysis/Figures/Upset_bothsamples_norep.pdf", height = 5.5, width = 5)
upset(fromList(upset_list),
              order.by = "freq",
              nsets = length(upset_list),
              sets.bar.color = c(
                  `Haplotype_0` = "#7BCEC8",
                  `Haplotype_7` = "#C49C94",
                  `Haplotype_4` = "#5DA069",
                  `Haplotype_1` = "#FF9896",
                  `Haplotype_9` ="#FFB14D",
                  `Haplotype_5` = "#6291EA",
                  `Haplotype_6` = "#C5B0D5",
                  `Haplotype_3` = "#17BECF",
                  `Haplotype_2` ="#B4E877",
                  #`Haplotype_12` = "#FFDC73",
                  #`Haplotype_10` = "#DE9ED6",
                  `Haplotype_11` = "#6B6ECF",
                  `Haplotype_8` = "#D46A97"
                  ))
dev.off()

## haps appear in either
C0 = seqtab_annotated %>%  filter(seq_id == "Haplotype_0")
C1 =seqtab_annotated %>% filter(seq_id == "Haplotype_1")
C2 =seqtab_annotated%>%  filter(seq_id == "Haplotype_2")
C3 =seqtab_annotated%>%  filter(seq_id == "Haplotype_3")
C4 =seqtab_annotated%>%  filter(seq_id == "Haplotype_4")
C5 =seqtab_annotated%>% filter(seq_id == "Haplotype_5")
C6 =seqtab_annotated%>%  filter(seq_id == "Haplotype_6")
C7 =seqtab_annotated%>%  filter(seq_id == "Haplotype_7")
C8 =seqtab_annotated%>%  filter(seq_id == "Haplotype_8")
C9 =seqtab_annotated%>%  filter(seq_id == "Haplotype_9")
C10 =seqtab_annotated%>%  filter(seq_id == "Haplotype_10")
C11=seqtab_annotated%>%  filter(seq_id == "Haplotype_11")
C12=seqtab_annotated%>%  filter(seq_id == "Haplotype_12")


upset_list = list(Haplotype_0 =C0$sample_single %>% unique(),
                  Haplotype_1 = C1$sample_single %>% unique(),
                  Haplotype_2 = C2$sample_single %>% unique(),
                  Haplotype_3 = C3$sample_single %>% unique(),
                  Haplotype_4 = C4$sample_single %>% unique(),
                  Haplotype_5 = C5$sample_single %>% unique(),
                  Haplotype_6 = C6$sample_single %>% unique(),
                  Haplotype_7 = C7$sample_single %>% unique(),
                  Haplotype_8 = C8$sample_single %>% unique(),
                  Haplotype_9 = C9$sample_single %>% unique(),
                  Haplotype_10 = C10$sample_single %>% unique(),
                  #Haplotype_12 = C12$sample_single %>% unique(),
                  Haplotype_11 = C11$sample_single %>% unique()
                  )

pdf("../Amplicon_analysis/Figures/Upset_eithersamples.pdf", height = 5.5, width = 5)
upset(fromList(upset_list),
              order.by = "freq",
              nsets = length(upset_list),
              sets.bar.color = c(
                  `Haplotype_0` = "#7BCEC8", 
                  `Haplotype_6` = "#C5B0D5",
                  `Haplotype_7` = "#C49C94", 
                  `Haplotype_1` = "#FF9896", 
                  `Haplotype_4` = "#5DA069",
                  `Haplotype_9` ="#FFB14D",
                  `Haplotype_5` = "#6291EA",
                  `Haplotype_12` = "#FFDC73",
                  `Haplotype_3` = "#17BECF",
                  `Haplotype_8` = "#D46A97", 
                  `Haplotype_2` ="#B4E877", 
                  `Haplotype_11` = "#6B6ECF",
                  `Haplotype_10` = "#DE9ED6"
                  ))
dev.off()

```



# Mom vs baby sample at birth -NOMRMALIZED TO GROUP SIZE

figures 3a and 3b

How many times a halpotype appeared in a baby sample only/total number of baby samples, number of times it appeared in a mom sample only/total number of mom samples, and number of times it appeared in a baby and a mom sample/total number of samples
```{r}
fisher_results = data.frame(Cluster = character(),
                            pvalue = numeric(),
                            OR = numeric(),
                            num_birth = numeric(), num_mom = numeric(), num_both = numeric(), num_none = numeric())

Cluster = "Haplotype_0"
haplotype_list = c("Haplotype_0", "Haplotype_1",
                   #"Haplotype_2", 
                   "Haplotype_3","Haplotype_4", "Haplotype_5", "Haplotype_6", "Haplotype_7", "Haplotype_8", "Haplotype_9",
                   #"Haplotype_10",
                   "Haplotype_12")

total_baby = both_samples %>% filter(sample_type == "0m") %>% select(family) %>% unique()  %>% nrow 
total_mom = both_samples %>% filter(sample_type == "mother") %>% select(family) %>% unique()  %>% nrow 
total_samp = total_baby +total_mom

for (Cluster in haplotype_list %>% unique() ) {
  
 contigency_table = both_samples %>% 
  filter(total_reads >100) %>% 
  #select(-seq, -seq_id, -freq, -mother) %>% unique %>% 
  filter(sample_type %in% c("mother", "0m")) %>% 
  pivot_wider(id_cols = c("sample_single","family" , "sample_type"),
              names_from = seq_id,
              values_from = merged_freq,
              values_fn = sum) %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% ####!! Need to expand out the matrix to include 0s for clusters not expressed
  pivot_longer(cols =haplotype_list, names_to = "seq_id", values_to = "freq") %>% 
  pivot_wider(id_cols = c(seq_id, family), names_from = sample_type, values_from = freq, values_fn = sum ) %>% 
  na_if(0)%>% 
  filter(seq_id == Cluster) %>% 
        mutate(samp_birth = ifelse( !is.na(.[3]), 1,0),
           samp_mom = ifelse(!is.na(.[4]), 1, 0),
           samp_both = ifelse(samp_birth == 1 & samp_mom == 1, 1, 0),
           samp_none = ifelse(samp_birth == 0 & samp_mom ==0, 1, 0)) %>% 
      mutate(num_birth = sum(samp_birth),
           num_mom = sum(samp_mom),
           num_both = sum(samp_both),
           num_none = sum(samp_none)) %>% select(num_birth, num_mom, num_both, num_none) %>% unique
  
  contingency_2b2 <- matrix(c(contigency_table$num_both,
                              contigency_table$num_birth,
                              contigency_table$num_mom,
                              contigency_table$num_none),nrow=2,ncol=2,dimnames=list(c("mom","noMom"), c("Birth","noBirth")))
  
    ftest = fisher.test(contingency_2b2,alternative = "less")
    
    fisher_results  %>% add_row("Cluster" = Cluster,
                              "pvalue" = ftest$p.value,
                              "OR" = ftest$estimate,
                              "num_birth"= contigency_table$num_birth/total_baby, 
                              "num_mom"= contigency_table$num_mom/total_mom, 
                              "num_both" = contigency_table$num_both/total_samp, 
                              "num_none" =contigency_table$num_none) -> fisher_results
}


fisher_results = fisher_results %>% 
  mutate(p_adjust = p.adjust(pvalue, method = "BH"))


fisher_results %>%
  mutate(BirthvsMom = num_birth-num_mom) %>% 
  ggplot() +
  geom_hline(yintercept = 0)+
  geom_bar(aes(y = BirthvsMom, x = Cluster, fill = Cluster), stat = "identity")+
  scale_fill_brewer(palette="Set3")+
  #scale_fill_manual(values = group.colors)+
  theme(legend.position = "none", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
fisher_results %>%
  ggplot() +
  geom_point(aes(y = num_both, x = Cluster), color ="#7C4A8A", size = 4)+
  scale_fill_brewer(palette="Set3")+
  #scale_fill_manual(values = group.colors)+
  geom_point(aes(y = num_mom, x = Cluster), color = "#E25D8D", size = 4, shape = 15)+
  geom_point(aes(y = num_birth, x = Cluster), color ="#27A4C3", size = 4, shape = 17)+
  theme_bw()+
  theme(legend.position = "bottom",  axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ 
  ylab("Number of Samples, normalized to sample pool")+xlab("Haplotypes")

fisher_results %>%
  pivot_longer(cols = c("num_birth", "num_mom", "num_both"), names_to = "MomOrBaby", values_to = "sample_num") %>% 
  mutate(Samples_found = case_when(MomOrBaby =="num_birth"~"0m", 
                                   MomOrBaby =="num_mom" ~ "mother",
                                   TRUE~"both")) %>% 
  filter(Samples_found != "both") %>% 
  ggplot() +
  geom_col(aes(fill  = Samples_found, y = sample_num, x = Cluster), position = "dodge", width = .5, color = "black")+
  scale_fill_manual(values = sample.colors)+
  theme(legend.position = "top",  axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Proportion of Samples with Haplotypes")+xlab("Haplotypes") +
  theme(  axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

fisher_results %>%
  pivot_longer(cols = c("num_birth", "num_mom", "num_both"), names_to = "MomOrBaby", values_to = "sample_num") %>% 
  mutate(Samples_found = case_when(MomOrBaby =="num_birth"~"0m", 
                                   MomOrBaby =="num_mom" ~ "mother",
                                   TRUE~"both")) %>% 
  filter(Samples_found != "both") %>% 
  mutate(Cluster = ordered(Cluster, 
                           levels = c("Haplotype_4", "Haplotype_9", "Haplotype_1","Haplotype_12","Haplotype_11", "Haplotype_5", "Haplotype_10","Haplotype_6", "Haplotype_0", "Haplotype_3", "Haplotype_8", "Haplotype_7","Haplotype_2"))) %>% 
  ggplot() +
  geom_col(aes(fill  = Samples_found, x = sample_num, y = Cluster), position = "dodge", width = .5, color = "black")+
  scale_fill_manual(values = sample.colors)+
  xlab("Proportion of Samples with Haplotypes")+ylab("Haplotypes") 


```


